---
title: "AMR-16S"
author: "JM"
date: "2023-02-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load libraries

```{r, echo=FALSE}
library(dada2)
library(ggplot2)
library(phyloseq)
library(vegan)
library(knitr)
library(ALDEx2)
library(CoDaSeq)
library(ANCOMBC)
library(dplyr)
library(nloptr)
library(tibble)
library(randomcoloR)
writeLines(capture.output(sessionInfo()), "sessionInfo.txt")
```

## Quality-filter the sequencing reads and create Amplicon Sequence Variant (ASV) tables with DADA2

Put unjoined R1 and R2 fastq files, with adaptors and primers previously removed with cutadapt into a directory for DADA2. Here, our forward and reverse fastq filenames have format: SAMPLENAME_R1_cut.fastq.gz and SAMPLENAME_R2_cut.fastq.gz


```{r, echo=FALSE}
path <- "~/Documents/SWS6323_MiniWorkshop/cutadapt"
list.files(path)
fnFs <- sort(list.files(path, pattern="_R1_cut.fastq.gz", full.names = TRUE))
fnRs <- sort(list.files(path, pattern="_R2_cut.fastq.gz", full.names = TRUE))
sample.names <- sapply(strsplit(basename(fnFs), "_"), `[`, 1)
# Perform filtering and trimming
filt_path <- file.path(path, "filtered") 
filtFs <- file.path(filt_path, paste0(sample.names, "_F_filt.fastq.gz"))
filtRs <- file.path(filt_path, paste0(sample.names, "_R_filt.fastq.gz"))
out <- filterAndTrim(fnFs, filtFs, fnRs, filtRs, truncLen=c(150,150),
                     maxN=0, maxEE=c(2,2), truncQ=2, rm.phix=TRUE,
                     compress=TRUE, multithread=TRUE)
head(out)
# Learn the Error Rates, it TAKES TIME!
errF <- learnErrors(filtFs, multithread=TRUE)
errR <- learnErrors(filtRs, multithread=TRUE)
plotErrors(errF, nominalQ=TRUE)
# Dereplicate the filtered fastq files
derepFs <- derepFastq(filtFs, verbose=TRUE)
derepRs <- derepFastq(filtRs, verbose=TRUE)
names(derepFs) <- sample.names
names(derepRs) <- sample.names
# Infer the sequence variants in each sample
dadaFs <- dada(derepFs, err=errF, multithread=TRUE)
dadaRs <- dada(derepRs, err=errR, multithread=TRUE)
# Inspecting the dada-class object returned by dada:
dadaFs[[1]]
# Merge the denoised forward and reverse reads:
mergers <- mergePairs(dadaFs, derepFs, dadaRs, derepRs, verbose=TRUE)
# Inspect the merger data.frame from the first sample
head(mergers[[1]])
# Construct sequence table
seqtab <- makeSequenceTable(mergers)
dim(seqtab)
#Remove chimeric sequences:
seqtab.nochim <- removeBimeraDenovo(seqtab, method="consensus", multithread=TRUE, verbose=TRUE)
dim(seqtab.nochim)
sum(seqtab.nochim)/sum(seqtab)
# Inspect distribution of sequence lengths
table(nchar(getSequences(seqtab)))
# Track reads through the pipeline
getN <- function(x) sum(getUniques(x))
track <- cbind(out, sapply(dadaFs, getN), sapply(mergers, getN), rowSums(seqtab))
colnames(track) <- c("input", "filtered", "denoised", "merged", "tabled")
rownames(track) <- sample.names
head(track)
write.table(track, "dada_read_stats.txt",sep="\t",col.names=NA)
saveRDS(seqtab, "~/Documents/SWS6323_MiniWorkshop/seqtab.rds") 
```

## Assign taxonomy in DADA2

Make sure the taxonomy reference database is in your working directory. Keep the database file gzipped. Adjust path name below. This step is very time consuming.

When taxonomy assignment is complete, we will use base R and phyloseq to clean up the taxonomy table. First, we will replace NAs and empty cells with the lowest taxonomy classification available. Second, we will use phyloseq to remove reads that are classified as Eukaryotes or unclassified at the domain level (ie, we are keeping only Bacteria and Archaea because that is what our primers target).

```{r, echo=FALSE}
taxa <- assignTaxonomy(seqtab.nochim, "~/Documents/Github/silva_nr99_v138.1_train_set.fa.gz", multithread=TRUE)
# FIX the NAs in the taxa table
taxon <- as.data.frame(taxa,stringsAsFactors=FALSE)
taxon$Phylum[is.na(taxon$Phylum)] <- taxon$Kingdom[is.na(taxon$Phylum)]
taxon$Class[is.na(taxon$Class)] <- taxon$Phylum[is.na(taxon$Class)]
taxon$Order[is.na(taxon$Order)] <- taxon$Class[is.na(taxon$Order)]
taxon$Family[is.na(taxon$Family)] <- taxon$Order[is.na(taxon$Family)]
taxon$Genus[is.na(taxon$Genus)] <- taxon$Family[is.na(taxon$Genus)]
write.table(taxon,"silva_taxa_table.txt",sep="\t",col.names=NA)
write.table(seqtab.nochim, "silva_otu_table.txt",sep="\t",col.names=NA)

# Create phyloseq object from otu and taxonomy tables from dada2, along with the sample metadata.
otu <- read.table("silva_otu_table.txt",sep="\t",header=TRUE, row.names=1)
taxon <- read.table("silva_taxa_table.txt",sep="\t",header=TRUE,row.names=1)
samples<-read.table("metadata.txt",sep="\t",header=T,row.names=1)
OTU = otu_table(otu, taxa_are_rows=FALSE)
taxon<-as.matrix(taxon)
TAX = tax_table(taxon)
sampledata = sample_data(samples)
ps <- phyloseq(otu_table(otu, taxa_are_rows=FALSE), 
               sample_data(samples), 
               tax_table(taxon))
ps 

# remove chloroplasts and mitochondria and Eukaryota
get_taxa_unique(ps, "Family")
get_taxa_unique(ps, "Order")
get_taxa_unique(ps, "Kingdom")
ps <- subset_taxa(ps, Family !="Mitochondria")
ps <- subset_taxa(ps, Order !="Chloroplast")
ps <- subset_taxa(ps, Kingdom !="Eukaryota")
ps <- subset_taxa(ps, Kingdom !="NA")
get_taxa_unique(ps, "Family")
get_taxa_unique(ps, "Order") 
get_taxa_unique(ps, "Kingdom")
ps  #6138 taxa and 52 samples

# Now export cleaned otu and taxa tables from phyloseq for future reference
otu = as(otu_table(ps), "matrix")
taxon = as(tax_table(ps), "matrix")
metadata = as(sample_data(ps), "matrix")
write.table(otu,"silva_nochloronomito_otu_table.txt",sep="\t",col.names=NA)
write.table(taxon,"silva_nochloronomito_taxa_table.txt",sep="\t",col.names=NA)

# export ASV table as relative abundance
ps_ra<-transform_sample_counts(ps, function(OTU) OTU/sum(OTU))
otu_ra = as(otu_table(ps_ra), "matrix")
write.table(otu_ra,"silva_nochloronomito_otu_table_RA.txt",sep="\t",col.names=NA)

```


# Now, time to explore the data.

```{r, echo=FALSE}
# load in data and create phyloseq object
otu <- read.table("silva_nochloronomito_otu_table.txt",sep="\t",header=TRUE, row.names=1)
taxon <- read.table("silva_nochloronomito_taxa_table.txt",sep="\t",header=TRUE,row.names=1)
samples<-read.table("metadata.txt",sep="\t",header=T,row.names=1)
OTU = otu_table(otu, taxa_are_rows=FALSE)
taxon<-as.matrix(taxon)
TAX = tax_table(taxon)
sampledata = sample_data(samples)
ps <- phyloseq(otu_table(otu, taxa_are_rows=FALSE), 
               sample_data(samples), 
               tax_table(taxon))
ps # 6138 taxa and 52 samples
psnb = subset_samples(ps, colony != "blank") #remove sample blanks
psnb # 6138 taxa and 50 samples
# remove disease samples to compare before and after amox treatment
ps1 = subset_samples(psnb, type != "disease") 
ps1 # 6138 taxa and 35 samples
otu = as(otu_table(ps1), "matrix")
taxon = as(tax_table(ps1), "matrix")
metadata = as(sample_data(ps1), "matrix")
write.table(otu,"silva_nochloronomito_otu_table_ps1.txt",sep="\t",col.names=NA)
write.table(taxon,"silva_nochloronomito_taxa_table_ps1.txt",sep="\t",col.names=NA)
write.table(metadata,"metadata_ps1.txt",sep="\t",col.names=NA)

ps5<-filter_taxa(ps1, function(x) mean(x) >5, TRUE)
ntaxa(ps5) #112
ps10<-filter_taxa(ps1, function(x) mean(x) >10, TRUE)
ntaxa(ps10) #78
get_taxa_unique(ps1, "Genus") #887
get_taxa_unique(ps5, "Genus") #61
get_taxa_unique(ps10, "Genus") #46

# Now export filtered otu and taxa tables from phyloseq for future reference
otu_ps5 = as(otu_table(ps5), "matrix")
taxon_ps5 = as(tax_table(ps5), "matrix")
metadata = as(sample_data(ps5), "matrix")
write.table(otu_ps5,"silva_nochloronomito_otu_table_ps5.txt",sep="\t",col.names=NA)
write.table(taxon_ps5,"silva_nochloronomito_taxa_table_ps5.txt",sep="\t",col.names=NA)
write.table(metadata,"metadata_ps5.txt",sep="\t",col.names=NA) 

#relative abundance
ps5_ra<-transform_sample_counts(ps5, function(OTU) OTU/sum(OTU))
otu_ps5_ra = as(otu_table(ps5_ra), "matrix")
write.table(otu_ps5_ra,"silva_nochloronomito_otu_table_ps5_RA.txt",sep="\t",col.names=NA)

```

# Perform center-log-ratio transformation on ASVs and calculate Aitchison Distance and principal components

```{r, echo=FALSE}
# load in data and create phyloseq object
otu <- read.table("silva_nochloronomito_otu_table_ps5.txt",sep="\t",header=TRUE, row.names=1)
taxon <- read.table("silva_nochloronomito_taxa_table_ps5.txt",sep="\t",header=TRUE,row.names=1)
samples<-read.table("metadata_ps5.txt",sep="\t",header=T,row.names=1)
OTU = otu_table(otu, taxa_are_rows=FALSE)
taxon<-as.matrix(taxon)
TAX = tax_table(taxon)
sampledata = sample_data(samples)
ps <- phyloseq(otu_table(otu, taxa_are_rows=FALSE), 
               sample_data(samples), 
               tax_table(taxon))
ps #112 taxa and 35 samples

# First, replace 0 values with an estimate (because normalization is taking log, can't have 0)
# Also transposing here, need samples as rows
d.czm <- cmultRepl(t(otu), method="CZM", label=0, z.warning=1)
# Perform the center-log-ratio (CLR) transformation 
d.clr <- codaSeq.clr(d.czm)
# transpose matrix of CLR transformed data for ordination and dendrogram
E.clr <- t(d.clr)
# plot compositional PCA biplot (perform a singular value decomposition)
d.pcx <- prcomp(E.clr)
# calculate percent variance explained for the axis labels
pc1 <- round(d.pcx$sdev[1]^2/sum(d.pcx$sdev^2),2)
pc2 <- round(d.pcx$sdev[2]^2/sum(d.pcx$sdev^2),2)
xlab <- paste("PC1: ", pc1, sep="")
ylab <- paste("PC2: ", pc2, sep="")
#biplot(d.pcx, cex=c(0.6,0.4), var.axes=F,scale=1, xlab=xlab, ylab=ylab)
summary(d.pcx)
str(d.pcx)
screeplot(d.pcx)
# Make a pretty PCA plot with ggplot
df_out <- as.data.frame(d.pcx$x)
theme_set(theme_bw()+theme(panel.grid.major = element_blank(),panel.grid.minor = element_blank()))
cols<-c("before"="#6CD5D9","after"="#107A86")
samples$type<-factor(samples$type, levels=c("before","after"))
pdf("PCA.pdf",bg ="white",width=8.5)
p<-ggplot(df_out,aes(x=PC1,y=PC2,fill=samples$type,shape=samples$fraction))
p<-p+geom_point(size=3)+
  theme(axis.title = element_text(size=14))+
  theme(axis.text=element_text(size=12))+
  theme(legend.title = element_text(size=14))+
  theme(legend.text = element_text(size=12))+
  scale_shape_manual(values=c(21,22))+
  scale_fill_manual(values=cols)+
  guides(fill = guide_legend(override.aes=list(shape=21)))+
  facet_grid(~samples$fraction)
p + labs(x=xlab, y=ylab, fill="Type",shape="Fraction") +coord_fixed()
dev.off()

####### Use phyloseq/vegan to perform PERMANOVA
# set metadata as factors
type<-as.character(samples$type)
frac<-as.character(samples$fraction)
# permanova between groups using Aitchison distance
dist.clr <- dist(E.clr)
perm<-adonis2(dist.clr~type*frac,as(sample_data(ps),"data.frame"))
print(perm)
```


Stacked bar charts
```{r, echo=FALSE}
# load in data and create phyloseq object
otu <- read.table("silva_nochloronomito_otu_table_ps5.txt",sep="\t",header=TRUE, row.names=1)
taxon <- read.table("silva_nochloronomito_taxa_table_ps5.txt",sep="\t",header=TRUE,row.names=1)
samples<-read.table("metadata_ps5.txt",sep="\t",header=T,row.names=1)
OTU = otu_table(otu, taxa_are_rows=FALSE)
taxon<-as.matrix(taxon)
TAX = tax_table(taxon)
sampledata = sample_data(samples)
ps <- phyloseq(otu_table(otu, taxa_are_rows=FALSE), 
               sample_data(samples), 
               tax_table(taxon))
ps #112 taxa and 35 samples

ps_ra<-transform_sample_counts(ps, function(OTU) OTU/sum(OTU))
get_taxa_unique(ps_ra, "Class") #19
get_taxa_unique(ps_ra, "Order") #37
get_taxa_unique(ps_ra, "Family") #53
get_taxa_unique(ps_ra, "Genus") #61

sample_data(ps_ra)$type<-factor(sample_data(ps_ra)$type, levels=c("before","after"))

n <- 19
# after plotting, you can re-run the next line to create a different selection of colors
palette <- distinctColorPalette(n)

pdf("barchart_Class.pdf",width=11)
p1=plot_bar(ps_ra, "colony" ,fill="Class")+
  facet_grid(fraction~type,scales="free",space="free")+
  geom_bar(aes(fill=Class), stat="identity",position="stack")+
  theme_bw()+
  theme(strip.text=element_text(face="bold", size=12))+
  theme(axis.text.x=element_text(size=12))+
  theme(axis.text.y=element_text(size=12))+
  scale_fill_manual(values=palette)+
  #theme(axis.title.x = element_blank())+
  theme(legend.position = "bottom")
p1
dev.off()



n <- 37
palette <- distinctColorPalette(n)

pdf("barchart_Order.pdf",width=11)
p1=plot_bar(ps_ra, "colony" ,fill="Order")+
  facet_grid(fraction~type,scales="free",space="free")+
  geom_bar(aes(fill=Order), stat="identity",position="stack")+
  theme_bw()+
  theme(strip.text=element_text(face="bold", size=12))+
  theme(axis.text.x=element_text(size=12))+
  theme(axis.text.y=element_text(size=12))+
  scale_fill_manual(values=palette)+
  #theme(axis.title.x = element_blank())+
  theme(legend.position = "bottom")
p1
dev.off()


n <- 53
palette <- distinctColorPalette(n)

pdf("barchart_Family.pdf",width=11)
p1=plot_bar(ps_ra, "colony" ,fill="Family")+
  facet_grid(fraction~type,scales="free",space="free")+
  geom_bar(aes(fill=Family), stat="identity",position="stack")+
  theme_bw()+
  theme(strip.text=element_text(face="bold", size=12))+
  theme(axis.text.x=element_text(size=12))+
  theme(axis.text.y=element_text(size=12))+
  scale_fill_manual(values=palette)+
  #theme(axis.title.x = element_blank())+
  theme(legend.position = "bottom")
p1
dev.off()

n <- 61
palette <- distinctColorPalette(n)

pdf("barchart_Genus.pdf",width=11)
p1=plot_bar(ps_ra, "colony" ,fill="Genus")+
  facet_grid(fraction~type,scales="free",space="free")+
  geom_bar(aes(fill=Genus), stat="identity",position="stack")+
  theme_bw()+
  theme(strip.text=element_text(face="bold", size=12))+
  theme(axis.text.x=element_text(size=12))+
  theme(axis.text.y=element_text(size=12))+
  scale_fill_manual(values=palette)+
  #theme(axis.title.x = element_blank())+
  theme(legend.position = "bottom")
p1
dev.off()


```










ANCOM-BC
https://bioconductor.org/packages/release/bioc/vignettes/ANCOMBC/inst/doc/ANCOMBC2.html


```{r, echo=FALSE}
# Start by creating phyloseq object using full dataset (low abundance ASVs are not removed).

otu <- read.table("silva_nochloronomito_otu_table_ps1.txt",sep="\t",header=TRUE, row.names=1)
taxon <- read.table("silva_nochloronomito_taxa_table_ps1.txt",sep="\t",header=TRUE,row.names=1)
samples<-read.table("metadata_ps1.txt",sep="\t",header=T,row.names=1)
OTU = otu_table(otu, taxa_are_rows=FALSE)
taxon<-as.matrix(taxon)
TAX = tax_table(taxon)
sampledata = sample_data(samples)
ps1 <- phyloseq(otu_table(otu, taxa_are_rows=FALSE), 
               sample_data(samples), 
               tax_table(taxon))
ps1 #6138 taxa and 35 samples

set.seed(123)
output = ancombc2(data = ps1, assay_name = "counts", tax_level = "Genus",
                  fix_formula = "type", rand_formula = NULL,
                  p_adj_method = "holm", pseudo = 0, pseudo_sens = TRUE,
                  prv_cut = 0.10, lib_cut = 1000, s0_perc = 0.05,
                  group = "type", struc_zero = FALSE, neg_lb = FALSE,
                  alpha = 0.05, n_cl = 2, verbose = TRUE,
                  global = FALSE, pairwise = FALSE, 
                  dunnet = FALSE, trend = FALSE,
                  iter_control = list(tol = 1e-5, max_iter = 20, 
                                      verbose = FALSE),
                  em_control = list(tol = 1e-5, max_iter = 100),
                  lme_control = NULL, mdfdr_control = NULL, 
                  trend_control = NULL)

res_prim = output$res
output$res
write.table(res_prim,file="ancomBC-results-BH-genus-by-treatment.txt")
tab_sens = output$pseudo_sens_tab


```


